## Scope and severity

Shortly after [version 1.0](http://dx.doi.org/10.5281/zenodo.7930001) 
of this code was published, IUCN has 
[clarified](https://www.iucnredlist.org/resources/threat-classification-scheme)
that _severity_ should be understood as the population decline 
"within the scope of the particular threat". 
Previously, the definition of severity had been ambiguous, and the Norwegian
Red Lists were based on an understanding of severity as the decline of the 
_entire population_. Whereas the current IUCN definition requires that 
threat scores are estimated from the product of scope and severity, 
the alternative definition required scope to be ignored.

Version 1.1 of this code now includes the possibility to choose between
these two definitions. (This choice is handled by the variable 
`useIUCNthreats`, see [species.md](species.md).)

The multiplicative approach had earlier been applied by Garnett et al. 
([2018](http://dx.doi.org/10.1111/cobi.13220), appendix S3) 
and Mair et al. ([2021](http://dx.doi.org/10.1038/s41559-021-01432-0), 
supplementary table 2).
It should be noted, however, that our code does not use the same threat weights
as these earlier studies. The main reasons for this are:

* Garnett et al. and Mair et al. assumed uniform distributions within each
scope and severity class. We do the same for scopes (which are fractions of
population sizes) but not for severities (which are population declines).
Especially in the highest severity class ("very rapid decline" according to
IUCN, "rapid decline" according to the Norwegian Red Lists), which spans the
interval up to and including 100% decline (i.e. extinction), it seems
unrealistic to treat extinction and a 30% decline as equiprobable. We therefore
assumed a decreasing distribution for the highest severity interval (and an
increasing distribution for the lowest of the Norwegian severity intervals,
which is "negligible"; however, this does not make sense for IUCN's lowest
severity interval, which is "no decline").
* Garnett. et al. and Mair et al. used the averages of the products of the limits
of each interval, whereas we use the products of the averages of the limits of
each interval.

The differences are here illustrated with the IUCN classes for scope and severity:

**(1)** By using _averages of products_ of the limits of each interval, 
the earlier studies obtained the following values 
(avoiding their rounding errors, however):
```{r}
ScopeSev <- list(
  c("whole_population", "majority", "minority"),
  c("very_rapid", "rapid", "slow", "negligible", "no_decline")
)
average <- function(x, y) (x + y) / 2
table <- average(ScopeIUCN$lower[3:1] %*% t(SeverityIUCN$lower[5:1]),
                 ScopeIUCN$upper[3:1] %*% t(SeverityIUCN$upper[5:1]))
print(round(matrix(table, 3, 5, dimnames = ScopeSev), 3))
```

**(2)** By using _products of averages_ of the limits of each interval instead, 
one obtains:
```{r}
table <-   average(   ScopeIUCN$lower[3:1],    ScopeIUCN$upper[3:1]) %*%
         t(average(SeverityIUCN$lower[5:1], SeverityIUCN$upper[5:1]))
print(round(matrix(table, 3, 5, dimnames = ScopeSev), 3))
```
For some combinations of scope and severity, the differences may be negligible. 
In the lower left corner of the matrix, the differences are clear, however.

Method (2) is statistically more meaningful than method (1). If the true values 
of _scope_ and _severity_ are uniformly distributed within their respective 
intervals, their product is not. Method (2) takes this into account, whereas 
method (1) assumes that the _products_ of the true values of scope and severity 
are uniformly distributed. The same analogously holds for other distributions: 
whichever distribution one assumes for scope and severity, their product will 
necessarily follow a different distribution.

**(3)** In addition to using products of averages, we use a decreasing distribution 
for the highest severity class. We therefore obtain:
```{r}
table <-   average(   ScopeIUCN$lower[3:1],       ScopeIUCN$upper[3:1]) %*%
         t(average(SeverityIUCN$lower[5:1], c(0.65, 0.3, 0.2, 0.02, 0)))
# That's a shortcut which reproduces our mean for the highest severity class
print(round(matrix(table, 3, 5, dimnames = ScopeSev), 3))
```


