---
title: Threats to species
subtitle: "Quantification of threats on the Norwegian Red Lists of species"
author: "Hanno Sandvik"
date: "17 April 2023"
output:
  md_document:
    toc: yes
---

This R code can be used to run the analyses of the Norwegian Red Lists for species 
described in the paper "Metrics for quantifying the contributions of 
different threats to Red Lists of species and ecosystems"



## Variables

The following variables can be used to adjust the output.

**(1) Name of the data file.** If you haven't saved the data file to 
your root directory, the file path needs to be included in the file name.
```{r}
file <- "species.csv"
```

**(2) Number of simulations.** 
NB: the default takes several hours!
For exploration purposes, `nsim <- 1000` will suffice.
For pure illustration, `nsim <- 100` is enough.
```{r}
nsim <- 100000
```

**(3) Re-create published estimates?** 
Decides whether the estimation of confidence intervals
should be re-created exactly as published (if TRUE) or be based on 
novel random numbers (if FALSE)
```{r}
re.create <- TRUE
```

**(4) Handling of DD species.** 
Decides whether Data Deficient species are excluded (if FALSE)
or randomly assigned to other Red List categories (if TRUE)
```{r}
includeDD <- TRUE
```

**(5) Handling of unknown threats.**
Decides whether (if TRUE) or not (if FALSE) unknown threat factors
should be inferred from the distribution of the known threat factors 
```{r}
inferThreats <- FALSE
```

**(6) Abbreviations used in the dataset.**
What are the abbreviations used for real population changes and for unknown threats?
(Defaults to the abbreviation used in the dataset analysed in the paper.
May need to be adjusted for other datasets.)
```{r}
realChange      <- "realpopu"
unknownThreat   <- "unknownf"
unknownTiming   <- "unknownt"
unknownScope    <- "unknownp"
unknownSeverity <- "unknownd"
```

**(7) Timings to include. **What is (are) the abbreviation(s) of the timing 
categories that should be considered (defaults to "ongoing").
```{r}
inclTiming <- "ongoingt"
```
If _all_ threats are to be included, irrespective of timing, this would need to be 
replaced (in terms of the abbreviations used in thus dataset) by
`inclTiming <- c("onlypast", "suspendd", "ongoing", "onlyfutu", "unknownt")`

**(8) File names of figures.** If you want to display the figures on screen,
keep the default. If you want to create PNG file, specify the file names
(including paths).
```{r}
fig1  <- ""
fig2  <- ""
fig3  <- ""
figS1 <- ""
figS2 <- ""
figS3 <- ""
```

**(9) Column names.** 
Column names in the dataset which contain Red List Categories, threat factors,
reasons for change in category, and generation time, respectively. 
The three former ones need to be followed by the year of assessment
(for change, the year of the _second_ of the two relevant assessments).
So if the column name containing Red List Categories is _not_ named
something like "Categ21" or "Categ2021", this needs to be adjusted here!
```{r}
Categ  <- "Categ"
Threat <- "Threat"
Change <- "Change"
GTime  <- "GenTime"
```



## Constants 

Constants should not normally need to be changed.
Changing them entails modifying some underlying assumptions.

**(1) Red List Categories** and their weights, extinction probabilities etc.
This data frame needs to contain all Red List Categories used in the 
Red List analysed of species or ecosystems that have been evaluated.

* The column "LC" identifies the Red List Category "Least Concern" (defaults to IUCN's abbreviation).
* The column "EX" identified the Red List Categories for extinction (defaults to IUCN IUCN's abbreviations).
* The column "wt" provides the Red List Weight of the Category (defaults to equal-steps weighting).
* The columns "lowP" and "uppP" provide the lower and upper threshold values for extinction probability (defaults to IUCN Red ListCriterion E).
* The columns "lowT" and "uppT" provide the lower and upper threshold values for extinction time frames in _years_ (defaults to IUCN Red List Criterion E).
* The columns "lowG" and "uppG" provide the lower and upper threshold values for extinction time in _generations_ (defaults to IUCN Red List Criterion E).
* The column "distr" provides the distribution of extinction probabilities within the interval.
* The column "beta" is not currently needed (but may be needed if "distr" is changed).
```{r}
RLcateg <- data.frame(
  name  = c(  "LC",   "NT",   "VU",   "EN",   "CR",   "RE",   "EW",   "EX"),
  LC    = c(  TRUE,  FALSE,  FALSE,  FALSE,  FALSE,  FALSE,  FALSE,  FALSE),
  EX    = c( FALSE,  FALSE,  FALSE,  FALSE,  FALSE,   TRUE,   TRUE,   TRUE),
  wt    = c(     0,      1,      2,      3,      4,      5,      5,      5),
  lowP  = c(  0.00,   0.05,   0.10,   0.20,   0.50,   1.00,   1.00,   1.00),
  uppP  = c(  0.00,   0.10,   0.20,   0.50,   1.00,   1.00,   1.00,   1.00),
  lowT  = c(   100,    100,    100,     20,     10,     10,     10,     10),
  uppT  = c(   100,    100,     20,     10,     10,     10,     10,     10),
  lowG  = c(     0,      0,      0,      5,      3,      1,      1,      1),
  uppG  = c(     0,      0,      5,      3,      1,      1,      1,      1),
  distr = c("unif", "unif", "unif", "unif", "decr", "unif", "unif", "unif"),
  beta  = c(    NA,     NA,     NA,     NA,     NA,     NA,     NA,     NA),
  stringsAsFactors = FALSE
)
```

**(2) Data deficiency**. What is the abbreviation used for the "Data Deficient" 
Red List Category? (Defaults to IUCN's abbreviation.)
```{r}
DD <- "DD"
```

**(3) Not Evaluated.** Which Red List Categories exist for species or ecosystems 
that have _not_ been evaluated? (Defaults to IUCN's abbreviations for the Red List 
Categories "Not Applicable" and "Not Evaluated".)
```{r}
notEval <- c("NA", "NE")
```

**(4) Downlisting.** What is added to a Red List Category to indicate downlisting?
(Defaults to the degree symbol.)
```{r}
downlistSymbol <- "°"
```

**(5) Severities** and their threshold values.
This data frame needs to contain all severity categories of threats used in the 
Red List analysed.
The data frame defaults to the severity categories used in Norwegian Red Lists,
where values correspond to the declines in population size over 10 years or
3 generations (whichever is largest) caused by a threat.

* The column "name" contains the abbreviations used for the severity categories.
* The column "lower" contains the lower limit of the respective interval.
* The column "upper" contains the upper limit of the respective interval.
* The column "distr" contains the distribution of values within the respective 
interval (possible values: "unif", "incr", "decr", "beta").
* The column "beta" contains the beta parameter of a Beta distribution
(a numeric values if `distr == "beta"`, and `NA` otherwise).
```{r}
Severity <- data.frame(
  name  = c("negldecl", "slowdecl", "rapidecl", "unknownd"),
  lower = c(      0.00,       0.02,       0.20,       0.00),
  upper = c(      0.02,       0.20,       1.00,       1.00),
  distr = c(    "incr",     "unif",     "decr",     "beta"),
  beta  = c(        NA,         NA,         NA,         20),
  stringsAsFactors = FALSE
)
```

**(6) Time frame** for the expected loss of species or ecosystems, in years
(defaults to 50 years).
```{r}
TimeFrame <- 50
```



## Preliminaries

Load the set of functions belonging to this depository:
```{r}
eval(parse(text = readLines("function.R")))
```

Define further required variables, 
based on the variables and constants specified above:
```{r}
LC      <- RLcateg$name[RLcateg$LC]
extinct <- RLcateg$name[RLcateg$EX]
LC.EX   <- RLcateg$name
RedListCat <- c(LC.EX, DD, notEval)
```



## Read and check the data

Read the dataset "Norwegian Red List for species":
```{r}
RL <- read.csv2(file, as.is=TRUE, dec=".", na.strings="n/a", encoding="latin1")
```

Check whether the data are as expected:
```{r}
{
  usedCategories <- checkRL(RL)
  years <- identifyYears(RL)
  cat("\nYears included in this dataset:\n")
  print(years)
  threats <- identifyThreats(RL)
  cat("\nThreats reported in this dataset:\n")
  print(threats)
}
```

Ensure that `RedListCat` and `LC.EX` only contain categories 
that are actually used:
```{r}
RedListCat <- RedListCat %A% usedCategories
LC.EX      <-      LC.EX %A% usedCategories
```



## Prepare the data frame for analyses

**(1) Reverse uplisting**:
```{r}
RL <- uplist(RL) 
```

**(2) Back-cast** knowledge from the most recent Red List to earlier ones:
```{r}
RL <- backCast(RL)
```

**(3) Calculate extinction probabilites** for all species:
```{r}
RL <- calcLoss(RL)
```

**(4) Add columns for all threat factors**:
```{r}
RL <- addThreats(RL)
```



## Summarise the data

Calculate RLIs for the three Red Lists 
(corrected for knowledge in the most recent one):
```{r}
{
  RLI21 <- RLI(RL$Categ21.21)
  RLI15 <- RLI(RL$Categ15.21)
  RLI10 <- RLI(RL$Categ10.21)
  cat("Red List Index 2010: " %+% RLI10 %+% "\n")
  cat("Red List Index 2015: " %+% RLI15 %+% "\n")
  cat("Red List Index 2021: " %+% RLI21 %+% "\n")
}
```

Create a list to summarise the Red Lists for 2010, 2015 and 2021:
```{r}
Table3 <- matrix(as.numeric(NA), 9, length(RedListCat) + 3, dimnames=list(
  "RL" %+% c("2010°", "2010", "2010(15)", "2010(21)",
             "2015°", "2015", "2015(21)", "2021°", "2021"),
  c("N", RedListCat, "RLI", "Cum.ELS50")
))
alphabetic <- sort(RedListCat)
```

Data for Red Lists 2010 and 2015 (prior to back-casting) have to be added
manually, because they cannot be re-created fully from the 2021 Red List data.
Source for these data:

* [Artsdatabanken (2010)](http://www.artsportalen.artsdatabanken.no/)
* [Artsdatabanken (2015)](https://www.artsdatabanken.no/Rodlista2015) 
```{r}
Table3[1, match(alphabetic, colnames(Table3))] <-
  c(284, 809, 890, 16762, 2580, 6528, 1310, 127, 1265)
Table3[2, match(alphabetic, colnames(Table3))] <-
  c(290, 809, 908, 16745,   NA,   NA, 1302, 127, 1266)
Table3[5, match(alphabetic, colnames(Table3))] <-
  c(247, 755, 901, 17594, 3018, 6095, 1302, 119, 1294)
Table3[6, match(alphabetic, colnames(Table3))] <- 
  c(252, 755, 916, 17579,   NA,   NA, 1297, 119, 1294)
tb <- table(substr(RL$Categ21, 1, 2))
Table3[8, match(names(tb), colnames(Table3))] <- tb
Loss21o <- LoS(substr(RL$Categ21, 1, 2), RL$GenTime)
```

Summarise the Red Lists:
```{r}
tab <- summariseRL(RL)
Table3[c(4, 7, 9), colnames(tab)] <- tab[c(3, 5, 6), ]
Table3[3, ] <- Table3[4, ] - Table3[7, ] + Table3[6, ]
Table3[, "N"] <- apply(Table3[, RedListCat], 1, sum, na.rm=T)
Table3[, "RLI"] <- 1 - apply(t(Table3[,LC.EX]) * 0:5, 2, sum, na.rm=T) / 
  apply(Table3[, LC.EX], 1, sum, na.rm=T) / max(RLcateg$wt,   na.rm=T)
Table3 <- Table3[, !is.na(apply(Table3 > 0, 2, any))]
```

Calculate means per Red List Category
(needed to approximate species loss for data 
that are not based on the 2021 Red List):
```{r}
mn10 <- mn15 <- rep(0, length(RedListCat))
names(mn10) <- names(mn15) <- RedListCat
for (i in RedListCat) {
  mn10[i] <- mean(RL$Loss10.21[which(RL$Categ10.21 == i)])
  mn15[i] <- mean(RL$Loss15.21[which(RL$Categ15.21 == i)])
}
Table3[1, "Cum.ELS50"] <- sum(mn10 * Table3[1, RedListCat], na.rm=T)
Table3[2, "Cum.ELS50"] <- sum(mn10 * Table3[2, RedListCat], na.rm=T)
Table3[3, "Cum.ELS50"] <- sum(mn15 * Table3[3, RedListCat], na.rm=T)
Table3[5, "Cum.ELS50"] <- sum(mn15 * Table3[5, RedListCat], na.rm=T)
Table3[6, "Cum.ELS50"] <- sum(mn15 * Table3[6, RedListCat], na.rm=T)
Loss21o[which(isDD(RL$Categ21))] <- RL$Loss21[which(isDD(RL$Categ21))]
Table3[8, "Cum.ELS50"] <- sum(Loss21o, na.rm=T)
rm(alphabetic, tb, tab, mn10, mn15, Loss21o)
```

Print the table (which underlies Table 3 of the paper):
```{r}
print(Table3)
```



## Analysis of threat factors

Estimate DeltaRLI:
```{r}
DRLI <- DeltaRLI(RL)
print(DRLI)
```

Estimate deltaRLI and ELS50:
```{r}
drli <- dRLI(RL)
print(drli)
```

Confidence intervals on RLI:
```{r, eval=FALSE, echo=TRUE}
print(confidenceRLI(RL, nsim, "Categ21"))
```

Confidence intervals on DeltaRLI, deltaRLI and ELS50:
```{r, eval=FALSE, echo=TRUE}
results <- simulateDRLI(RL, nsim)
```



## Figures

### Figure 1
The following script recreates Fig 1 only if the
above default parameter values are left unchanged.

Simplify the table by collapsing minor threats:
```{r}
DRLI. <- DRLI
DRLI.[10, ] <- apply(DRLI.[c(1, 2, 5, 7, 9, 10, 11, 13), ], 2, sum)
DRLI. <- DRLI.[c(6, 10, 3, 12, 8, 4), ]
DRLI. <- DRLI.[, c(1:3, 3)]
DRLI. <- rbind(0, DRLI.)
DRLI. <- rbind(DRLI., 0)
```

Plot a graph for DeltaRLI:
```{r}
if (nchar(fig1)) {
 png(fig1, 1500, 1200, res = 180)
}
par(mai = c(0.96, 0.96, 0.12, 0.06), family = "sans") # xlim = c(2009.5, 2027.5)
plot(0, 0, xlim = c(2009.5, 2029.5), ylim = c(0.9198, 0.92022),
  xaxs = "i", yaxs = "i", xaxt = "n", yaxt = "n", xlab = "", 
  ylab = "Red List Index",
  bty = "n", cex.axis = 1.2, cex.lab = 1.8)
axis(1, c(2009, 2021.5),              F, T, tcl = 0,        lwd = 1.5, lend = 1)
axis(1, 2009:2021,                    F, T,                 lwd = 1.5, lend = 1)
axis(1, c(2010, 2015, 2021),          T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, seq(0.9198, 0.9202, 0.00010), T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, seq(0.9198, 0.9202, 0.00001), F, T, tcl = -0.25,    lwd = 1.5, lend = 1)
mtext("Year of Red List assessment", 1, 3, F, 2015.5, cex = 1.8)
#x <- c(2010, 2015, 2021, 2023)
x <- c(2010, 2015, 2021, 2022)
#DRLI.[, 4] <- DRLI.[, 3] + c(0, 0, 0, 0, 0, 0, -0.00000887, 0)
DRLI.[, 4] <- DRLI.[, 3] + c(0, 0, 0, 0, 0, 0, -0.00000724, 0)
lines(x[4:1], rep(RLI10, 4), lty = "12",     lwd = 9.6, col = grey(0.84))
lines(x[1:4], c(RLI10, RLI15, RLI21, RLI21), lwd = 9.6, col = grey(0.84))
for (i in 2:7) {
  lines(x, RLI10 + DRLI.[i, ], lty = i - 1, lwd = 2.4)
  points(x[2:3], RLI10 + DRLI.[i, 2:3], pch = c(1, 4, 21, 24, 22, 25, 23)[i],
    cex = 1.8, bg = "black", lwd = 2.4, ljoin = 1)
  #text(2023, RLI10 + DRLI.[i, 4],
  text(2022, RLI10 + DRLI.[i, 4],
    c("", "land-use change", "other/unknown", "climate change", "pollution",
      "native species", "disturbance")[i] %+% " (" %+%
      c(rep("+", 3), rep("", 5))[i] %+% 
      gsub("-", "−", formatC(DRLI.[i, 3], 6, format="f")) %+% ")",
    pos = 4, cex = 1.2)
}
text(2022, RLI21, "RLI (+0.000096)", pos = 4, cex = 1.2)
text(2022, RLI10, "no change (+0.000000)", pos = 4, cex = 1.2)
text(2015.5, 0.92021, expression(bold(Delta*RLI)), cex = 1.8)
if (nchar(fig1)) {
  dev.off()
}
```


### Figure S1
The following script recreates Fig S1 only if `inferThreats == TRUE`
and the remaining default parameter values are is left unchanged.

Simplify the table by collapsing minor threats:
```{r}
DRLI. <- DRLI
DRLI.[10, ] <- apply(DRLI.[c(1, 2, 5, 7, 9, 10, 11, 13), ], 2, sum)
DRLI. <- DRLI.[c(6, 10, 3, 12, 8, 4), ]
DRLI. <- DRLI.[, c(1:3, 3)]
DRLI. <- rbind(0, DRLI.)
DRLI. <- rbind(DRLI., 0)
```

Plot a graph for DeltaRLI:
```{r}
if (nchar(figS1)) {
 png(figS1, 1500, 1200, res = 180)
}
par(mai = c(0.96, 0.96, 0.12, 0.06), family = "sans") # ylim=c(0.9199, 0.92015)
plot(0, 0, xlim = c(2009.5, 2027.5), ylim = c(0.9198, 0.92033),
  xaxs = "i", yaxs = "i", xaxt = "n", yaxt = "n", xlab = "", 
  ylab = "Red List Index",
  bty = "n", cex.axis = 1.2, cex.lab = 1.8)
axis(1, c(2009, 2021.5),               F, T, tcl = 0,        lwd = 1.5, lend = 1)
axis(1, 2009:2021,                     F, T,                 lwd = 1.5, lend = 1)
axis(1, c(2010, 2015, 2021),           T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, seq(0.9198, 0.9203, 0.00010), T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, seq(0.9198, 0.9203, 0.00001), F, T, tcl = -0.25,    lwd = 1.5, lend = 1)
mtext("Year of Red List assessment", 1, 3, F, 2015.5, cex = 1.8)
x <- c(2010, 2015, 2021, 2023)
DRLI.[,4] <- DRLI.[,3] + c(0, 0, 0, 0, 0, 0, -0.00000876, 0)
lines(x[3:1], rep(RLI10, 3), lty = "12", lwd = 9.6, col = grey(0.84))
lines(x[1:3], c(RLI10, RLI15, RLI21),    lwd = 9.6, col = grey(0.84))
for (i in 2:7) {
  lines(x, RLI10 + DRLI.[i, ], lty = i - 1, lwd = 2.4)
  points(x[2:3], RLI10 + DRLI.[i, 2:3], pch=c(1, 4, 21, 24, 22, 25, 23)[i],
    cex = 1.8, bg = "black", lwd = 2.4, ljoin = 1)
  text(2023, RLI10 + DRLI.[i, 4],
    c("", "land-use change", "other/unknown", "climate change", "pollution",
      "native species", "disturbance")[i],
    pos = 4, cex = 1.2)
}
text(2015.5, 0.92032, expression(bold(Delta*RLI)), cex = 1.8)
if (nchar(figS1)) {
  dev.off()
}
```


### Figure 2
The following script recreates Fig 2 only if the
above default parameter values are left unchanged.

Simplify the table by collapsing minor threats:
```{r}
drli. <- drli$dRLI
ELS. <- drli$ELS50
for (i in 1:3) {
  drli.["otherthr", i] <- sum(drli.[c("otherthr", "unknownf", "alienspe",
    "huntgath", "outsiden", "natcatas", "bycatchc", "nothreat"), i])
  ELS. ["otherthr", i] <- sum(ELS. [c("otherthr", "unknownf", "alienspe",
    "huntgath", "outsiden", "natcatas", "bycatchc", "nothreat"), i])
}
drli. <- drli.[-which(rownames(drli.) %in%
  c("unknownf","alienspe","huntgath","outsiden","natcatas","bycatchc","nothreat")),]
ELS.  <- ELS. [-which(rownames(ELS.)  %in%
  c("unknownf","alienspe","huntgath","outsiden","natcatas","bycatchc","nothreat")),]
ELS.  <- ELS. [order(drli.[, 3], decreasing=T),]
ELS.  <- ELS. [, c(1:3, 3)]
drli. <- drli.[order(drli.[, 3], decreasing=T),]
drli. <- drli.[, 3:1]
drli. <- rbind(0, drli.)
drli. <- rbind(0, drli.)
```

Plot a graph for deltaRLI:
```{r}
if (nchar(fig2)) {
 png(fig2, 1500, 1200, res=180)
}
par(mai=c(0.96, 0.96, 0.06, 0.06), family="sans")
# plot(0, 0, xlim=c(2009.5, 2027.5), ylim=c(0.91, 1.002),
plot(0, 0, xlim=c(2009.5, 2029.5), ylim=c(0.91, 1.008),
  xaxs="i", yaxs="i", xaxt="n", yaxt="n", xlab="", ylab="Red List Index",
  bty="n", cex.axis=1.2, cex.lab=1.8)
axis(1, c(2009, 2021.5),     F, T, tcl=0,        lwd=1.5, lend=1)
axis(1, 2009:2021,           F, T,               lwd=1.5, lend=1)
axis(1, c(2010, 2015, 2021), T, T, cex.axis=1.2, lwd=1.5, lend=1)
axis(2, seq(0.8, 1, 0.01),   F, T,               lwd=1.5, lend=1)
axis(2, seq(0.8, 1, 0.02),   T, T, cex.axis=1.2, lwd=1.5, lend=1)
mtext("Year of Red List assessment", 1, 3, F, 2015.5, cex=1.8)
x <- c(2021, 2015, 2010)
for (i in 3:nrow(drli.)) {
  polygon(c(x, rev(x)),
          1 - c(apply(drli.[1:(i-1),], 2, sum), rev(apply(drli.[1:i,], 2, sum))),
          border=NA, col=grey(1.32 - i * 0.12))
}
y0 <- rep(RLI21, 3)
y2 <- 1 - (sum(drli.[1:8,1]) + sum(drli.[1:7,1])) / 2
#y2 <- RLI21 + 0.005
for (i in 7:3) {
  y1 <- 1 - apply(drli.[1:i,], 2, sum)
  lines(x, y1, lwd=2.4)
#  lines(c(2021, 2023), c(mean(c(y0[1], y1[1])), y2), lwd=1.8)
  lines(c(2021, 2022), c(mean(c(y0[1], y1[1])), y2), lwd=1.8)
#   text(2023, y2, 
  text(2022, y2, 
    c("", "land-use change", "other/unknown", "climate change",
      "pollution", "native species", "disturbance")[i] %+%
      " (" %+% formatC(drli.[i + 1, 1], 4, format="f") %+% ")",
    pos=4, cex=1.2)
  y0 <- y1
  y2 <- y2 + 0.005
}
#lines(c(2021, 2023), rep(y2, 2), lwd=1.8)
lines(c(2021, 2022), rep(y2, 2), lwd=1.8)
#text(2023, y2, "land-use change", pos=4, cex=1.2)
text(2022, y2, "land-use change" %+%
               " (" %+% formatC(drli.[3, 1], 4, format="f") %+% ")",
     pos=4, cex=1.2)
lines(x, rep(1, 3), lwd=4.8)
lines(c(2021, 2022), rep(1, 2), lwd=1.8)
text(2022, 1, "reference value", font=2, pos=4, cex=1.2)
lines(x, c(RLI21, RLI15, RLI10), lwd=4.8)
lines(c(2021, 2022), c(RLI21, y2 - 0.03), lwd=1.8)
text(2022, y2 - 0.03, "RLI", font=2, pos=4, cex=1.2)
text(2015.5, 1.005, expression(bold(delta*RLI)), cex = 1.8)
if (nchar(fig2)) {
  dev.off()
}
```


### Figure 3
The following script recreates Fig 3 only if the
above default parameter values are left unchanged.

Plot a graph for ELS50:
```{r}
if (nchar(fig3)) {
 png(fig3, 1500, 1200, res = 180)
}
par(mai = c(0.96, 0.96, 0.06, 0.06), family = "sans")
# plot(0, 0, xlim=c(2009.5, 2027.5), ylim=c(lg(8), 3), 
plot(0, 0, xlim=c(2009.5, 2029.5), ylim=c(lg(8), 3.1), 
     xaxs = "i", yaxs = "i", xaxt = "n", yaxt = "n", xlab = "", 
     ylab = "Expected loss of species",
     bty = "n", cex.axis = 1.2, cex.lab = 1.8)
axis(1, c(2009, 2021.5),          F, T, tcl = 0,        lwd = 1.5, lend = 1)
axis(1, 2009:2021,                F, T,                 lwd = 1.5, lend = 1)
axis(1, c(2010, 2015, 2021),      T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, 0:3, c(1, 10, 100, 1000), T,    cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, lg(c(2:9, seq(20, 90, 10), seq(200, 900, 100))),
                                  F, T, tcl = -0.25,    lwd = 1.5, lend = 1)
mtext("Year of Red List assessment", 1, 3, F, 2015.5, cex = 1.8)
# x <- c(2010, 2015, 2021, 2023)
x <- c(2010, 2015, 2021, 2022)
ELS.[3, 4] <- 10^(1/3 * lg(ELS.[4, 4]) + 2/3 * lg(ELS.[6, 4]))
ELS.[5, 4] <- 10^(2/3 * lg(ELS.[4, 4]) + 1/3 * lg(ELS.[6, 4]))
for (i in 1:6) {
  lines (x,      lg(ELS.[i, ]), lty = i, lwd = 2.4)
  points(x[1:3], lg(ELS.[i, 1:3]), pch = c(4, 21, 24, 22, 25, 23)[i],
    cex = 1.8, bg = "black", lwd = 2.4, ljoin = 1)
# text(2023, lg(ELS.[i, 4]), 
  text(2022, lg(ELS.[i, 4]), 
    c("land-use change", "other/unknown", "climate change",
      "pollution", "native species", "disturbance")[i] %+%
      " (" %+% round(ELS.[i, 3]) %+% ")",
    pos = 4, cex = 1.2)
}
text(2015.5, 3.01, expression(bold(ELS[50])), cex = 1.8)
if (nchar(fig3)) {
  dev.off()
}
```


### Figure S2
The following script recreates Fig S2 only if `inferThreats == TRUE`
and the remaining default parameter values are left unchanged.

Simplify the table by collapsing minor threats:
```{r}
drli. <- drli$dRLI
ELS. <- drli$ELS50
for (i in 1:3) {
  drli.["otherthr", i] <- sum(drli.[c("otherthr", "unknownf", "alienspe", 
    "huntgath", "outsiden", "natcatas", "bycatchc", "nothreat"), i])
  ELS. ["otherthr", i] <- sum(ELS. [c("otherthr", "unknownf", "alienspe",
    "huntgath", "outsiden", "natcatas", "bycatchc", "nothreat"), i])
}
drli. <- drli.[-which(rownames(drli.) %in%
  c("unknownf", "alienspe", "huntgath", "outsiden", 
    "natcatas", "bycatchc", "nothreat")),]
ELS.  <- ELS. [-which(rownames(ELS.)  %in%
  c("unknownf", "alienspe", "huntgath", "outsiden",
    "natcatas", "bycatchc", "nothreat")),]
ELS. <-  ELS.[c("landusec", "otherthr", "climatec", 
                "pollutio", "nativesp", "disturba"), ]
ELS.  <- ELS. [, c(1:3, 3)]
drli. <- drli.[order(drli.[, 3], decreasing=T), ]
drli. <- drli.[, 3:1]
drli. <- rbind(0, drli.)
drli. <- rbind(0, drli.)
```

Plot a graph for deltaRLI:
```{r}
if (nchar(figS2)) {
 png(figS2, 1500, 1200, res = 180)
}
par(mai = c(0.96, 0.96, 0.06, 0.06), family = "sans")
plot(0, 0, xlim = c(2009.5, 2027.5), ylim = c(0.91, 1.008),
  xaxs = "i", yaxs = "i", xaxt = "n", yaxt = "n", xlab = "", 
  ylab = "Red List Index",
  bty = "n", cex.axis = 1.2, cex.lab = 1.8)
axis(1, c(2009, 2021.5),     F, T, tcl = 0,        lwd = 1.5, lend = 1)
axis(1, 2009:2021,           F, T,                 lwd = 1.5, lend = 1)
axis(1, c(2010, 2015, 2021), T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, seq(0.8, 1, 0.01),   F, T,                 lwd = 1.5, lend = 1)
axis(2, seq(0.8, 1, 0.02),   T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
mtext("Year of Red List assessment", 1, 3, F, 2015.5, cex = 1.8)
x <- c(2021, 2015, 2010)
for (i in 3:nrow(drli.)) {
  polygon(c(x, rev(x)),
          1 - c(apply(drli.[1:(i-1), ], 2, sum), rev(apply(drli.[1:i, ], 2, sum))),
          border = NA, col = grey(1.32 - i * 0.12))
}
y0 <- rep(RLI21, 3)
y2 <- 1 - (sum(drli.[1:8, 1]) + sum(drli.[1:7, 1])) / 2
for (i in 7:3) {
  y1 <- 1 - apply(drli.[1:i, ], 2, sum)
  lines(x, y1, lwd = 2.4)
  lines(c(2021, 2023), c(mean(c(y0[1], y1[1])), y2), lwd = 1.8)
  text(2023, y2, 
    c("", "land-use change", "other/unknown", "climate change",
      "pollution", "native species", "disturbance")[i],
    pos = 4, cex = 1.2)
  y0 <- y1
  y2 <- y2 + 0.005
}
lines(c(2021, 2023), rep(y2, 2), lwd = 1.8)
text(2023, y2, "land-use change", pos = 4, cex = 1.2)
lines(x, rep(1,3), lwd = 4.8)
lines(x, c(RLI21, RLI15, RLI10), lwd = 4.8)
text(2015.5, 1.005, expression(bold(delta*RLI)), cex = 1.8)
if (nchar(figS2)) {
  dev.off()
}
```


### Figure S3
The following script recreates Fig S3 only if `inferThreats == TRUE`
and the remaining default parameter values are left unchanged.

Plot a graph for ELS50:
```{r}
if (nchar(figS3)) {
 png(figS3, 1500, 1200, res = 180)
}
par(mai = c(0.96, 0.96, 0.06, 0.06), family = "sans")
plot(0, 0, xlim = c(2009.5, 2027.5), ylim = c(1, 3.2),
  xaxs = "i", yaxs = "i", xaxt = "n", yaxt = "n", xlab = "",
  ylab="Expected loss of species",
  bty = "n", cex.axis = 1.2, cex.lab = 1.8)
axis(1, c(2009, 2021.5),          F, T, tcl = 0,        lwd = 1.5, lend = 1)
axis(1, 2009:2021,                F, T,                 lwd = 1.5, lend = 1)
axis(1, c(2010, 2015, 2021),      T, T, cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, c(1, lg(1200)),           F, T, tcl = 0,        lwd = 1.5, lend = 1)
axis(2, 0:3, c(1, 10, 100, 1000), T,    cex.axis = 1.2, lwd = 1.5, lend = 1)
axis(2, lg(c(2:9, seq(20, 90, 10), seq(200, 900, 100))),
                                  F, T, tcl = -0.25,    lwd = 1.5, lend = 1)
mtext("Year of Red List assessment", 1, 3, F, 2015.5, cex = 1.8)
x <- c(2010, 2015, 2021, 2023)
ELS.[2, 4] <- 10^(3/4 * lg(ELS.[4, 4]) + 1/4 * lg(ELS.[6, 4]))
ELS.[5, 4] <- 10^(2/4 * lg(ELS.[4, 4]) + 2/4 * lg(ELS.[6, 4]))
ELS.[3, 4] <- 10^(1/4 * lg(ELS.[4, 4]) + 3/4 * lg(ELS.[6, 4]))
for (i in 1:6) {
  lines (x,      lg(ELS.[i, ]), lty = i, lwd = 2.4)
  points(x[1:3], lg(ELS.[i, 1:3]), pch = c(4, 21, 24, 22, 25, 23)[i],
    cex = 1.8, bg = "black", lwd = 2.4, ljoin = 1)
  text(2023, lg(ELS.[i, 4]), 
    c("land-use change", "other/unknown", "climate change",
      "pollution", "native species", "disturbance")[i],
    pos = 4, cex = 1.2)
}
text(2015.5, 3.12, expression(bold(ELS[50])), cex = 1.8)
if (nchar(figS3)) {
  dev.off()
}
```

